### 2013-01-21 edupar-13/conclusion.tex
Copyedit pass on intro,conclusion,landslide,outcomes. Add outcomes sentence about BKL.

% Due to the open-ended design problems students solve in their kernels,
% Landslide imposes a moderate annotation burden in order to instrument a kernel for systematic testing.
% However, by the very nature of having enough spare time to volunteer, the selection of students we worked with to use Landslide was necessarily biased towards the ``top of the class''.
In our experience,
the annotation burden imposed on ``top of the class'' students by Landslide is modest.
Though our experiments with these students show promising results, we seek to reduce Landslide's annotation burden to make it useful for struggling students as well.
%
We believe some annotations could be automatically detected;
for example, rather than requiring students to declare if their kernel has a preemption disabling mechanism more elaborate than disabling interrupts, Landslide could detect when a timer interrupt doesn't induce a context switch.
Other annotations, such as \x{tell_landslide_thread_runnable()},
might be automated by requiring students to structure code in
certain ways with pre-determined function names and parameters
(Landslide already automatically instruments the memory allocator we provide).
However, \shortversion{}{prescribing} such templates
\shortversion{would constrain}{has the downside of constraining}
the project's open-ended design space.
We wish to better understand the tradeoffs between
making Landslide more automatic and forcing particular design decisions on students.
% As open-ended decisions of interface design are important to the course philosophy, we foresee a tradeoff: in some cases, fundamental design challenges may be the cost for increasing Landslide's usability to enable teaching more advanced concepts such as SMP.

>>>>

% Due to the open-ended design problems students solve in their kernels,
% Landslide imposes a moderate annotation burden in order to instrument a kernel for systematic testing.
% However, by the very nature of having enough spare time to volunteer, the selection of students we worked with to use Landslide was necessarily biased towards the ``top of the class''.
In our experience,
the annotation burden imposed on ``top of the class'' students by Landslide is modest.
Though our experiments with these students show promising results, we seek to reduce Landslide's annotation burden to make it useful for struggling students as well.
%
We believe some annotations could be automatically detected;
for example, rather than requiring students to declare if their kernel has a preemption-disabling mechanism more elaborate than disabling interrupts, Landslide could detect when a timer interrupt doesn't induce a context switch.
Other annotations, such as \x{tell_landslide_thread_runnable()},
might be automated by requiring students to structure code in
certain ways with pre-determined function names and parameters
(Landslide already automatically instruments the memory allocator we provide).
However, prescribing such templates
\shortversion{would constrain}{has the downside of constraining}
the project's open-ended design space.
We wish to better understand the tradeoffs between
making Landslide more automatic and preserving the open-endedness of important design decisions.
% As open-ended decisions of interface design are important to the course philosophy, we foresee a tradeoff: in some cases, fundamental design challenges may be the cost for increasing Landslide's usability to enable teaching more advanced concepts such as SMP.

### 2013-01-21 edupar-13/landslide.tex
clarify systematic vs stress sentence
Copyedit pass on intro,conclusion,landslide,outcomes. Add outcomes sentence about BKL.

Landslide is an experimental tool for helping kernel authors find and debug concurrency errors.
It is based on the technique of systematic exploration~\cite{verisoft}, a way of exploring the state space of different thread interleavings in a concurrent system.
It follows in the footsteps of related tools such as
\shortversion{
CHESS~\cite{chess},
}{
CHESS~\cite{chess}, dBug~\cite{dbug-ssv}, and DeMeter~\cite{demeter},
}
but is the first tool we know of to apply systematic exploration in a kernel development environment.

Conceptually,
Landslide is designed as follows.
Students annotate their code so Landslide knows
which kernel thread is currently running.
After one kernel thread has run for some time,
Landslide triggers artificial clock interrupts
% to force a new thread to interleave with the current one.
to force the scheduler to run a different thread.
When a test program finishes execution according
to one pattern of thread switches,
Landslide rewinds the kernel's state
and resumes the test according to a different thread interleaving.
After each instruction,
Landslide applies several bug-detection predicates to
the kernel's state to detect
illegal heap accesses,
deadlock, infinite loops, and panics.
\shortversion{}{
In theory,
by forcing a thread switch after every non-scheduler instruction,
Landslide could apply its bug-detection predicates to every reachable
execution state.
Because this would require prohibitive time to complete,
in practice Landslide uses a variety of techniques to
thread-switch less often and to avoid repeating
bug-equivalent execution paths.
}
%Landslide is implemented as a module for Simics, the simulator our students use as an execution environment for their kernels. Simics can load Landslide while booting a student's kernel, and calls into Landslide once per simulated instruction and once per simulated memory access.
%Landslide uses this information to update its internal representation of the kernel's state, which it in turn uses to decide how to control the kernel's execution.

@@

Systematic exploration, Landslide's mechanism for testing many different concurrent scenarios, uses the idea of an {\em execution tree} to enumerate all possible thread interleavings of a specific test. These interleavings are defined around {\em decision points}, which intuitively indicate points during execution at which a preemption could cause different behaviour to arise.
Searching with few decision points results in coarser-grained interleavings, faster test completion, and less likelihood of finding unexpected bugs; searching with more decision points results in the opposite.
The advantage of systematic testing is the ability to force execution sequences that might be rare in simple testing.
% Unlike simple testing, systematic execution reaches all execution states, so it does not overlook rare bugs.

@@

Of course, additional decision points are necessary to expose more-subtle race conditions.
% \shortversion
% {In Section~\ref{sec:instrument} we describe an interface for adding new
% decision points to expand the state space,
% and also an interface for restricting Landslide's attention to certain kernel components,
% which helps cope with the exponentially-growing nature of state spaces with many decision points.}
{We provide an interface for adding new decision points to expand the state space, described in Section~\ref{sec:decision}.
In order to cope with the exponentially-growing nature of state spaces with many decision points,
\shortversion{
we also provide an interface for making Landslide consider
}{
Landslide uses the Dynamic Partial Order Reduction algorithm~\cite{dpor} to identify and prune redundant thread interleavings, the specifics of which are beyond the scope of this educational work~\cite{landslide}.
We also provide an interface for helping Landslide further reduce the state space by considering
}
decision points from only certain components of the kernel, which we discuss in Section~\ref{sec:focusing}.}

@@

When Landslide identifies a bug, it outputs a
\shortversion
{{\em decision trace} (Figure~\ref{fig:trace}).}
{{\em decision trace}, an example of which is shown in Figure~\ref{fig:trace}.}
The trace reports what kind of bug was detected
\shortversion
{}
{(for uses-after-free, it also prints stack traces to show when the block was allocated and freed),}
and also reports each decision point in the current interleaving: which thread was running, a trace of its stack when it was switched away from, and the thread that Landslide caused to preempt it.
With this trace, the student can better understand the concurrent execution that exposed the bug.

Note that our use of the term ``race condition'' throughout this paper refers to concurrency errors in general, a category that includes data races, atomicity violations, and nondeterministic deadlocks.
In contrast with analyses that directly detect data races~\cite{tsan}, Landslide does not identify bugs at suspicious memory accesses, but rather detects failure conditions that can arise from many different types of errors.
\shortversion{}{
In general,
% with false-negative bug detection,
a kernel might contain a buggy behaviour that Landslide could overlook (i.e., a false negative). However (except in the case of heuristic infinite loop detection), when Landslide does identify a bug, the student can be sure that an error exists.
}

@@

\subsubsection{Required annotations}
Users annotate their kernels to inform Landslide of certain important concurrency events during execution. We provide a set of annotation functions, named with the prefix \x{tell_landslide}, for this purpose. The annotations denote when a thread runs \x{fork}, \x{sleep}, or \x{exit}, when a thread is added to or removed from the runqueue, and when a thread becomes blocked on a mutex.
Figure~\ref{fig:annotation} shows annotated code corresponding to Figure~\ref{fig:threadfork}.

>>>>

Landslide is an experimental tool for helping kernel authors find and debug concurrency errors.
It is based on the technique of systematic exploration~\cite{verisoft}, a way of exploring the state space of different thread interleavings in a concurrent system.
It follows in the footsteps of related tools such as
\shortversion{CHESS~\cite{chess},}
{CHESS~\cite{chess}, dBug~\cite{dbug-ssv}, and DeMeter~\cite{demeter},}
but is the first tool we know of to apply systematic exploration in a kernel development environment.

Conceptually,
Landslide is designed as follows.
Students annotate their code so Landslide knows
which kernel thread is currently running.
After one kernel thread has run for some time,
Landslide triggers artificial clock interrupts
% to force a new thread to interleave with the current one.
to force the scheduler to run a different thread.
When a test program finishes execution according
to one pattern of thread switches,
Landslide rewinds the kernel's state
and resumes the test according to a different thread interleaving.
After each instruction,
Landslide applies several bug-detection predicates to
the kernel's state to detect
illegal heap accesses,
deadlock, infinite loops, and panics.
\shortversion{}{In theory,
by forcing a thread switch after every non-scheduler instruction,
Landslide could apply its bug-detection predicates to every reachable
execution state.
Because this would require prohibitive time to complete,
in practice Landslide uses a variety of techniques to
thread-switch less often and to avoid repeating
bug-equivalent execution paths.}
%Landslide is implemented as a module for Simics, the simulator our students use as an execution environment for their kernels. Simics can load Landslide while booting a student's kernel, and calls into Landslide once per simulated instruction and once per simulated memory access.
%Landslide uses this information to update its internal representation of the kernel's state, which it in turn uses to decide how to control the kernel's execution.

@@

Systematic exploration, Landslide's mechanism for testing many different concurrent scenarios, uses the idea of an {\em execution tree} to enumerate all possible thread interleavings of a specific test. These interleavings are defined around {\em decision points}, which intuitively indicate points during execution at which a preemption could cause different behaviour to arise.
Searching with few decision points results in coarser-grained interleavings, faster test completion, and less likelihood of finding unexpected bugs; searching with more decision points results in the opposite.
The primary advantage of systematic testing is the ability to force execution sequences that might be overlooked in simple stress testing.
% Unlike simple testing, systematic execution reaches all execution states, so it does not overlook rare bugs.

@@

Of course, additional decision points are often necessary to expose more-subtle race conditions.
% \shortversion
% {In Section~\ref{sec:instrument} we describe an interface for adding new
% decision points to expand the state space,
% and also an interface for restricting Landslide's attention to certain kernel components,
% which helps cope with the exponentially-growing nature of state spaces with many decision points.}
We provide an interface for adding new decision points to expand the state space.
%described in Section~\ref{sec:decision}.
In order to cope with the exponentially-growing nature of state spaces with many decision points,
\shortversion{we also provide an interface for making Landslide consider}
{Landslide uses the Dynamic Partial Order Reduction algorithm~\cite{dpor} to identify and prune redundant thread interleavings, the specifics of which are beyond the scope of this educational work~\cite{landslide}.
We also provide an interface for helping Landslide further reduce the state space by considering}
decision points from only certain components of the kernel.
%which we discuss in Section~\ref{sec:focusing}.
Section~\ref{sec:instrument} discusses these interface options.

@@

When Landslide identifies a bug, it outputs a
\shortversion
{{\em decision trace} (Figure~\ref{fig:trace}).}
{{\em decision trace}, an example of which is shown in Figure~\ref{fig:trace}.}
The trace reports what kind of bug was detected
\shortversion
{and}
{(for uses-after-free, it also prints stack traces to show when the block was allocated and freed), and}
also reports each decision point in the current interleaving: which thread was running, a trace of its stack when it was switched away from, and the thread that Landslide caused to preempt it.
With this trace, the student can better understand the concurrent execution that exposed the bug.

Note that our use of the term ``race condition'' throughout this paper refers to concurrency errors in general, a category that includes data races, atomicity violations, and nondeterministic deadlocks.
In contrast with analyses that directly detect data races~\cite{tsan}, Landslide does not identify bugs at suspicious memory accesses, but rather detects failure conditions that can arise from many different types of errors.
\shortversion{}{In general,
% with false-negative bug detection,
a kernel might contain a buggy behaviour that Landslide could overlook (i.e., a false negative). However (except in the case of heuristic infinite loop detection), when Landslide does identify a bug, the student can be sure that an error exists.
}

@@

\subsubsection{Required annotations}
Users annotate their kernels to inform Landslide of certain important concurrency events during execution. We provide a set of annotation functions, named with the prefix \x{tell_landslide}, for this purpose. The annotations denote when a thread runs \x{fork}, \x{sleep}, or \x{vanish}, when a thread is added to or removed from the runqueue, and when a thread becomes blocked on a mutex.
Figure~\ref{fig:annotation} shows annotated code corresponding to Figure~\ref{fig:threadfork}.

### 2013-01-21 edupar-13/outcomes.tex
Copyedit pass on intro,conclusion,landslide,outcomes. Add outcomes sentence about BKL.

In this section we briefly describe selected
outcomes of the Spring 2012 offering of 15-410.
We begin by discussing the results of the
pre-Landslide educational experience and then
report on the results of applying Landslide to
both kernels previously written by members of the course staff
and kernels written that semester by students.

@@

The SMP kernel-extension project
was a mixed success.
Of the twenty-four teams that attempted it,
one (solo) team was unable to submit a solution.
Nineteen of the remaining twenty-three groups
completed part of the project---at least enough to
boot multiple cores
into their scheduler.
Four of those nineteen groups were able to pass
almost all of our test suite,
which consisted of simple multi-threaded applications
and machines with various core counts\shortversion
{.}
{(one, two, or eight depending on the simulator configuration).}
Seven more groups were able to pass roughly half of the test suite.
Finally, it was the sense of the grading team that
most groups would have been able to substantially complete
the project given a further week of development time.
We view this outcome as an indication that feasible adjustments
to the course design would bring this material within reach
of most students.

>>>>

In this section we briefly describe selected
outcomes of the Spring 2012 offering of 15-410.
We begin by discussing the results of the
pre-Landslide educational experience and then
report on the results of applying Landslide both to
kernels previously written by members of the course staff
and to kernels written that semester by students.

@@

The SMP kernel-extension project
was a mixed success.
Of the twenty-four teams that attempted it,
one (solo) team was unable to submit a solution.
Nineteen of the remaining twenty-three groups
completed part of the project---at least enough to
boot multiple cores
into their scheduler.
Four of those nineteen groups were able to pass
almost all of our test suite,
which consisted of simple multi-threaded applications
and machines with various core counts\shortversion
{.}
{(one, two, or eight depending on the simulator configuration).}
Seven more groups were able to pass roughly half of the test suite.
None of the teams elected to use a ``big kernel lock'' to bypass any of the synchronization challenges.
Finally, it was the sense of the grading team that
most groups would have been able to substantially complete
the project given a further week of development time.
We view this outcome as an indication that feasible adjustments
to the course design would bring this material within reach
of most students.
